### 2024-03-08 docs/manuscript/manuscript.qmd
Ms minor Changes
update in discussion - mediating effects discussed - strenghts and limitations updated - some citations are still required [cita requerida]

**Background**: Physical fitness, body composition, and the autonomic response to exercise are important predictors of health-related outcomes in older adults, but their interplay is not fully understood, therefore, this study aims to assess the associations between these variables and possible mediating effects. **Methods**: an observational cross-sectional study was conducted in 81 community-dwelling older people, recruited by non-probabilistic sampling. To obtain body composition parameters, a bio-impedance analysis was made, then, the Short Physical Performance Battery (SPPB) was used to assess physical fitness, and the Heart Rate Variability (HRV) was recorded before, during and after the Two-Minute Step Test (TMST), to assess cardiac autonomic response to physical exercise. Geriatric depression and anxiety symptoms were assessed as secondary measures. Correlation and mediating effects analyses were carried out. **Results**: SPPB score was negatively related to body fat percentage (r = -0.273, p = 0.015), but positively related to SNS index during exercise (r = 0.313, p = 0.006). Also, a negative link existed between body fat and sympathetic activity: SNS index (r = -0.252, p = 0.030) and Stress index (r = 0.258, p = 0.027) during TMST. But a positive link with parasympathetic activity: RMSSD (r = 0.253, p = 0.029) and SDNN (r = 0.269, p = 0.020) **Conclusion**: the findings unveil a triangular relationship, where body fat percentage significantly impacts physical fitness and cardiac autonomic activity during exercise, while a greater sympathetic activity correlates with higher physical fitness. Additionally, physical fitness moderates the link between body fat and cardiac autonomic activity, emphasizing its importance for older adults' health. Further research is needed to explore psychological variables' moderating effects.

@@

Physical fitness has been classically defined as the ability to perform daily tasks with vigor and without undue fatigue, and with ample energy to enjoy leisure-time pursuits and to meet unforeseen emergencies [@pmid:3920716]. Therefor, it is a crucial component of healthy aging. The term encompasses a multitude of aspects, typically including cardiorespiratory endurance, muscular endurance and strength, body composition, and flexibility [@pmid:3920711]. Numerous studies have demonstrated the positive impact of physical fitness on overall health and longevity, highlighting its role in reducing the risk of chronic diseases, and improving quality of life and functional independence in older adults [@pmid:9177581; @pmid:29293447; @pmid:30768197; @pmid:36332759]

@@

Despite extensive research into physical fitness, body composition, and cardiovascular health in older adults, gaps persist in the understanding of their interrelationships and underlying mechanisms. One area of interest is the relationship between physical fitness, body fat percentage, and cardiac autonomic response to exercise in older individuals. Although previous studies have explored these factors independently, this study explore their collective influence on physical fitness in the aging population.

The conceptualization of this study is based on the recognition of the interplay between physical fitness, body composition, and autonomic cardiovascular regulation in aging. We aim to answer the following research question: How does physical fitness relate to body fat percentage and cardiac autonomic response to exercise in older people? By elucidating these correlations, we hope to gain valuable insights into the relationships that underlie age-related modifications in physical fitness and identify potential interventions to enhance health outcomes in older adults. Even more, we seek to understand the moderating effect of frequent psychological traits, such as geriatric depression and anxiety.

@@

The primary objective of this study is to investigate the relationship between physical fitness, body fat percentage, and cardiac autonomic response to exercise in older people. Even more, We aimed to assess the moderating effect of psychological variables on these relationships, with implications for cardiovascular health and functional capacity in aging.

@@

81 community-dwelling older adults were recruited and selected by non-probabilistic sampling, from the Region of Magallanes and Chilean Antartica, Chile. They were included if: 65 years age or older; residents of the Region of Magallanes and Chilean Antartica, Chile; and being able to understand the study aims and instructions. But they were excluded in case of: diagnosis of congenital heart disease; consumption of beta-blocker drugs, taking stimulant substances within 24 h before assessments session; motor or cognitive disability; inability to understand instructions or written content; or presence of pain during cardiac or physical assessments.

@@

Additionally, and in the case on sympathetic indicators, the effect of body fat on SNS index during exercise (ADE = -0.050, CI~95%~[-0.107, 0.01], p = 0.077) was influenced by SPPB score in similar way as with RMSSD, potentiating the original main observed effect (ACME = -0.016, CI~95%~[-0.049, 0], p = 0.041; Total effect = -0.066, CI~95%~[-0.125, -0.010], p = 0.016). Similar effects were observed with TMST steps in this regard (ACME = -0.015, CI~95%~[-0.042, 0], p = 0.050; Total effect = -0.066, CI~95%~[-0.125, -0.010], p = 0.016). No other sympathetic indices were notoriously modified in the presence of either SPPB score, TMST steps or psychological variables.

@@

@@

This study aimed to investigate the relationship between physical fitness, body fat percentage, and cardiac autonomic response to exercise in older people. Through correlational analyses, we aimed to elucidate the  interplay between these factors and their implications for cardiac autonomic function during physical efforts.

@@

Furthermore, greater body fat percentage was related to a reduced sympathetic drive and higher parasympathetic activity during the TMST, as hypothesized. Throughout the course of physical efforts, the ANS is expected to shift into a sympathetic state, ensuring an efficient response to greater metabolic demands [@pmid:33246163]. However, aged individuals may present an impaired ANS response to physical exercise [@ijerph20054456]. In that sense, our results support that body composition may play a pivotal role in that response, generating novel insights for potential interventions.

Nonetheless, the moderating effect of physical fitness related measures on the relationship proposed earlier, suggests that greater physical activity could potentially counteract the negative effects of a high body fat percentage on cardiac autonomic response to exercise, possibly, at the same extent as seen in leaner individuals. This result highlights the importance of promoting a physically active lifestyle and favor the maintenance of physical capacity, regardless of nutritional status [@pmid:33402165; @pmid:30002462].

Regarding the link between physical fitness and cardiac autonomic response to exercise, our results evidenced that individuals with a high sympathetic drive while exercising, had greater physical fitness. Additionally, previous research had demonstrated the utility of HRV and exercise-induced responses in those metrics as a potential marker of cardiorespiratory fitness in these individuals [@pmid:36054204], supporting the current hypothesis that a greater sympathetic drive facilitates performance and its related measures during moderate intensity physical activities. Even more, it has been previously demonstrated that physically active individuals have an enhanced autonomic response to exercise, playing a pivotal role strengthening this hypothesis [@pmid:28721030; @pmid:24799513].

@@

However, our results did not evidence a moderating effect of either GDS-30 or BAI scores, challenging our initial hypothesis. These finding suggests a nuanced interplay between psychological variables and physical/physiological measures in older people, underscoring the need for further research to elucidate underlying mechanisms and variables beyond depression and anxiety. While the lack of evidence suggesting any moderating effects does not diminish the significance of psychological factors in aging, it highlights the complexity of their influence.

In addition to the potential strengths of this cross-sectional study, it is not without limitations. First, our study design prevents us from making causal inferences about the direction of the relationships, highlighting the relevance of experimental and longitudinal research exploring causal relations between our variables and potential underlying mechanisms. Second, a relatively small sample was included, limiting generalizability and statistical power of the analysis. Furthermore, the sample consisted mainly of women (82.7%), limiting our ability to make sex comparisons and decreasing the representativeness of the sample. It should also be noted that the relationships between physical fitness and cardiac autonomic response to exercise were not made within the same instrument, due to the possible lack of SPPB to generate an adequate physical stress, and of TMST to inform general physical fitness of the participants. Further research could investigate the impact of cardiac autonomic response on performance during the same task, in this age group.

@@

The results highlighted a structural bidirectional relationship among variables. Body fat percentage was inversely correlated to both physical fitness and cardiac autonomic activity during exercise, with sympathetic drive associated with physical fitness in older people. Furthermore, physical fitness may mediate the effect of body fat on cardiac autonomic activity during exercise, highlighting its pivotal role in older people's health. However, further research is needed to elucidate moderating effects of psychological variables in this field.

>>>>

**Background**: Physical fitness, body composition, and the autonomic response to exercise are important predictors of health-related outcomes in older adults, but their interplay is not fully understood, therefore, this study aims to assess the associations between these variables and possible mediating effects. **Methods**: an observational cross-sectional study was conducted in 81 community-dwelling older people, recruited by non-probabilistic sampling. To obtain body composition parameters, a bio-impedance analysis was made, then, the Short Physical Performance Battery (SPPB) was used to assess physical fitness, and the Heart Rate Variability (HRV) was recorded before, during and after the Two-Minute Step Test (TMST), to assess cardiac autonomic response to physical exercise. Geriatric depression and anxiety symptoms were assessed as secondary measures. Correlation and mediating effects analyses were carried out. **Results**: SPPB score was negatively related to body fat percentage (r = -0.273, p = 0.015), but positively related to SNS index during exercise (r = 0.313, p = 0.006). Also, a negative link existed between body fat and sympathetic activity: SNS index (r = -0.252, p = 0.030) and Stress index (r = 0.258, p = 0.027) during TMST. But a positive link with parasympathetic activity: RMSSD (r = 0.253, p = 0.029) and SDNN (r = 0.269, p = 0.020) **Conclusion**: Body fat was inversely correlated to both physical fitness and cardiac autonomic activity during exercise, with sympathetic drive associated with physical fitness in older people. Furthermore, physical fitness may mediate the effect of body fat on cardiac autonomic activity during exercise, and total muscle mass may mitigate the negative effect of increased adiposity on physical fitness, highlighting its pivotal role in older people's health. However, further research is needed to elucidate mediating effects of psychological variables in this field.

@@

Physical fitness has been classically defined as the ability to perform daily tasks with vigor and without undue fatigue, and with ample energy to enjoy leisure-time pursuits and to meet unforeseen emergencies [@pmid:3920716]. Therefore, it is a crucial component of healthy aging. The term encompasses a multitude of aspects, typically including cardiorespiratory endurance, muscular endurance and strength, body composition, and flexibility [@pmid:3920711]. Numerous studies have demonstrated the positive impact of physical fitness on overall health and longevity, emphasizing its role in reducing the risk of chronic diseases, and improving quality of life and functional independence in older adults [@pmid:9177581; @pmid:29293447; @pmid:30768197; @pmid:36332759]

@@

Despite extensive research into physical fitness, body composition, and cardiovascular health in older adults, gaps persist in the understanding of their interrelationships and factors mediating them. One area of interest is the relationship between physical fitness, body fat percentage, and cardiac autonomic response to exercise in older individuals. This study explores their collective influence on physical fitness in the aging population.

The conceptualization of this investigation is based on the recognition of the interplay between physical fitness, body composition, and autonomic cardiovascular regulation in aging. We aim to answer the following research question: How does physical fitness relate to body fat percentage and cardiac autonomic response to exercise in older people? By elucidating these correlations, we hope to gain valuable insights into the relationships that underlie age-related modifications in physical fitness and identify potential interventions to enhance health outcomes in older adults. Even more, we seek to understand the mediating effect of the variables, including frequent psychological traits, such as geriatric depression and anxiety.

@@

The primary objective of this study is to investigate the relationship between physical fitness, body fat percentage, and cardiac autonomic response to exercise in older people. Even more, We aimed to assess the mediating effect of psychological variables on these relationships, with implications for cardiovascular health and functional capacity in aging.

@@

81 community-dwelling older adults were recruited and selected by non-probabilistic sampling, from the Region of Magallanes and Chilean Antartica, Chile. They were included if: 65 years age or older; residents of the Region of Magallanes and Chilean Antarctica, Chile; and being able to understand the study aims and instructions. But they were excluded in case of: diagnosis of congenital heart disease; consumption of beta-blocker drugs, taking stimulant substances within 24 h before assessments session; motor or cognitive disability; inability to understand instructions or written content; or presence of pain during cardiac or physical assessments.

@@

Additionally, and in the case on sympathetic indicators, the effect of body fat on SNS index during exercise (ADE = -0.050, CI~95%~[-0.107, 0.01], p = 0.077) was influenced by SPPB score in similar way as with RMSSD, enhancing the original main observed effect (ACME = -0.016, CI~95%~[-0.049, 0], p = 0.041; Total effect = -0.066, CI~95%~[-0.125, -0.010], p = 0.016). Similar effects were observed with TMST steps in this regard (ACME = -0.015, CI~95%~[-0.042, 0], p = 0.050; Total effect = -0.066, CI~95%~[-0.125, -0.010], p = 0.016). No other sympathetic indices were notoriously modified in the presence of either SPPB score, TMST steps or psychological variables.

@@

Main results are represented in the Figure X.

Fig. X. Diagram of main results.

****Insertar Figura resumen resultados****

Footnote: The figure represents the main results of this study. Blue arrows and boxes represent the correlation between variables, while orange arrows and boxes represent the factors mediating the above correlations. "+" means that the mediating factor enhances the main observed effect, contrary to "-", reflecting a diminished main effect when considering the mediating factor.

@@

This study aimed to investigate the relationship between physical fitness, body fat percentage, and cardiac autonomic response to exercise in older people. Through correlational and mediating analyses, we aimed to elucidate the interplay between these factors and their implications for cardiac autonomic function during physical efforts.

@@

However, the mediating analysis showed that total muscle mass diminished the main observed effect of body fat on physical fitness, with more muscled individuals displaying a better physical performance despite the body fat percentage. This finding suggests that older adults could be benefited of activities that promotes their muscle mass conservation, even when they have high body fat percentages [cita requerida]. Nevertheless, caution is advised in the interpretation of this result, as the close physiological association among increasing body fat and decreasing muscle mass during aging have been well-studied [cita requerida: sarcopenic obesity]

Regarding the link between physical fitness and cardiac autonomic response to exercise, our results evidenced that individuals with a high sympathetic drive while exercising, had greater physical fitness. Previous research had demonstrated the utility of HRV and exercise-induced responses in those metrics as a potential marker of cardiorespiratory fitness in these individuals [@pmid:36054204], supporting the current hypothesis that a greater sympathetic drive facilitates physical performance during moderate intensity activities. Even more, it has been previously demonstrated that physically active individuals have an enhanced autonomic response to exercise, playing a pivotal role and strengthening this hypothesis [@pmid:28721030; @pmid:24799513].

Furthermore, the increase in body fat percentage was related to a reduced sympathetic activity and a higher parasympathetic activity during TMST, as hypothesized (i.e., SNS index, SI, SDNN and RMSSD). Throughout the course of physical efforts, ANS is expected to shift into a sympathetic predominance, ensuring a sufficient response to higher metabolic demands [@pmid:33246163]. However, aged individuals may present an impaired ANS response to physical exercise [@ijerph20054456]. In that sense, our results support that body composition plays a role in autonomic response. Moreover, they suggest that the negative effects of body fat on physical fitness in older adults, may be partially attributable to less efficient autonomic inputs to the cardiovascular system [cita requerida].

Surprisingly, mediating analyses found that higher physical fitness (SPPB and TMST scores) intensifies the negative impact of body fat on autonomic control. This unexpected finding hints at a complex interplay between physical fitness, adiposity, and cardiac autonomic regulation. Our results indicate that in individuals with greater functional capacity, the influence of body fat on HRV is amplified rather than mitigated. This observation challenges traditional conceptions about the relationship between physical fitness and cardiac autonomic regulation, and raises questions about the underlying mechanisms involved.

As far as we know, there are not preceding studies addressing these findings. We posit that there exists further factors influencing the mediation effect. For instance, body distribution of the adipose tissue was not explored, specially when it is known that visceral fat has an important impact on cardiovascular health and its autonomic regulation [cita requerida]. Participants with higher fitness may have a different fat distribution compared with lower performance individuals, which may induce different effects on autonomic drive to the heart [cita requerida]. Furthermore, other physical physiological, demographic or psychological variables may be playing an unexplored role. However, these hypothesis must be examined under scientific standards. Therefore, exhaustive research is needed to address this intriguing interplay.

@@

However, our results did not evidence a mediating effect of either GDS-30 or BAI scores, challenging our initial hypothesis. These finding suggests a nuanced interplay between psychological variables and physical/physiological measures in older people, underscoring the need for further research to elucidate underlying mechanisms and variables beyond depression and anxiety. While the lack of evidence suggesting any mediating effects does not diminish the significance of psychological factors in aging, it highlights the complexity of their influence.

This study was strengthened by the fact that both SPPB and TMST measures consistently aligned in their interaction with other variables, which supports a global understanding of physical fitness in older people, rather than specific aspects, and by the exploration of mediating factors in the primary results. Nonetheless, this cross-sectional study, it is not without limitations. First, our study design prevents us from making causal inferences about the direction of the relationships, highlighting the relevance of experimental and longitudinal research exploring causal relations between our variables and potential underlying mechanisms. Second, a relatively small sample was included, limiting generalizability and statistical power of the analysis. Furthermore, the sample consisted mainly of women (82.7%), limiting our ability to make sex comparisons and decreasing the representativeness of the sample. Further research could investigate the impact of cardiac autonomic response on performance during the same task, in this age group.

@@

The results highlighted a structural bidirectional relationship among variables. Body fat percentage was inversely correlated to both physical fitness and cardiac autonomic activity during exercise, with sympathetic drive associated with physical fitness in older people. Furthermore, physical fitness may mediate the effect of body fat on cardiac autonomic activity during exercise, and total muscle mass may mitigate the negative effect of increased adiposity on physical fitness, highlighting its pivotal role in older people's health. However, further research is needed to elucidate mediating effects of psychological variables in this field.
